Extending K-Means Clustering with Ptolemy’s Inequality \thanks{This work is partly funded by DFG grant No. 454630593: EPIX – Efficient Ptolemaic Indexing.}

### 2024-11-15 review/rebuttal.md

Thank you for your reviews! Let me address your questions:

Total Runtime Measurements (R2, R3)
Our paper only considers the number of explicit distance evaluations to measure the efficiency of the algorithms we compare. Reviews 2 and 3 bring up the valid point of missing total execution time measurements. While we mention this fact as a limitation (in ch. 4.3, Limitations, point ii), our paper does not give a good explanation for this choice.

We chose to count distance evaluations because they are agnostic to the specific implementation of algorithms in machine code. The main disadvantage of reporting execution times is their strong dependence on the amount of optimization that went into the implementation.
For example, time measurements are affected by the programming language chosen, the amount of time spend on optimizing the source code, and the specific choice of hardware architecture (e.g. GPU or CPU).
Therefore, it takes considerable effort and care to make time measurements objective, reproducible, and generalizable. Consequently, in our short paper we focus on the theoretical, implementation-independent aspects of the discussed algorithms.

That said, the actual time measurements are important for real-world applications: Besides the distance evaluations, the caching and evaluation of the upper and lower bound approximations also takes time. While we reduce the total number of distance computations compared to Elkan's algorithm, our new bounds require more memory accesses and computations (cf. Eq. 3,4 and Eq. 5,6).
While this additional time is likely to be small compared to the time spent on distance evaluations, we plan to measure and report the actual ratio in a future full paper.

For the camera-ready version of our paper, we will include this elaboration and explicitly add execution time measurements as future work.

Discussion of Observed Behavior.
Review 3 notes that our discussion does not give a clear indication of how the chosen parameters relate to the speedups achieved, i.e. how the number of clusters $k$, the dataset, and its dimensionality influence the number of distance calculations saved. We agree that more precise rules predicting the actual speedups would be immensely useful.

However, we believe that we cannot, in good conscience, derive better rules from our data than those presented in the paper: While we did not cherry-pick the datasets used in our investigation, we support Review 3's notion that the speedups appear to be dataset-dependent (ch. 4.3, Limitations, point i). Similarly, we believe that a wider range of starting parameters (initializations, $k$s) is needed to draw more robust conclusions (ch. 4.3, Limitations, point iii).

Consequently, we feel confident that our discussion cannot go into more detail without running the significant risk of arriving at non-generalizable conclusions.

Analysis of the Set of Distance Computations.
Review 3 suggests an analysis of which distances are computed in each algorithm. We think this is an idea worth pursuing in the preparation of a full paper, but the actual implementation of this approach seems to require considerable effort.
While obtaining the data needed for such an analysis is straightforward, how to interpret it is still unclear to us at this time. Clusterings on data beyond the two-dimensional case are difficult to interpret visually, making it harder to develop meaningful measures for comparing the distance computation sets.

Minor Problems.
Review 3 correctly notes that Eq. 2 has a typo in the first summand of the RHS. We will correct this mistake in the camera-ready version. We will also improve the readability and accessibility of Fig. 1 by increasing its font size and using different line styles rather than relying on coloring alone.

Review 1 does not contain any questions or statements that need to be rebutted.

### 2024-11-17 review/rebuttal.md

Thank you for your reviews! We greatly appreciate your suggestions and will take them into consideration when updating our paper. Let me address your questions:

Total Runtime Measurements (R2, R3)
Our paper only considers the number of explicit distance evaluations to measure the efficiency of the algorithms we compare. Reviews 2 and 3 bring up the valid point of missing total execution time measurements. While we mention this fact as a limitation (ch. 4.3, Limitations, point ii), our paper does not give a good explanation for this choice.

We chose to count distance evaluations because they are agnostic to the specific implementation of algorithms in machine code. The main disadvantage of reporting execution times is their strong dependence on the amount of optimization that went into this implementation.
For example, time measurements are affected by the programming language chosen, the amount of time spend on optimizing the source code, and the specific choice of hardware architecture (e.g. GPU or CPU). It takes considerable effort and care to make time measurements objective, reproducible, and comparable. Consequently, in our short paper we focus on the theoretical, implementation-independent aspects of the discussed algorithms.

@@

That said, the actual time measurements are important for real-world applications: Besides the distance evaluations, the caching and evaluation of the upper and lower bound approximations also takes time. While we reduce the total number of distance computations compared to Elkan's algorithm, our new bounds require more memory accesses and computations (cf. Eq. 3,4 and Eq. 5,6).
While this additional time is likely to be small compared to the time spent on distance evaluations, we plan to measure and report the actual ratio in a future full paper.

Should the paper be accepted, we will include this elaboration and explicitly add execution time measurements as future work.

@@

However, we believe that we cannot, in good conscience, derive better rules from our data than those presented in the paper: While we did not cherry-pick the datasets used in our investigation, we support Review 3's notion that the speedups appear to be dataset-dependent (ch. 4.3, Limitations, point i). Similarly, we believe that a wider range of starting parameters (initializations, $k$s) is needed to draw more robust conclusions (ch. 4.3, Limitations, point iii).

Consequently, we feel confident that our discussion cannot go into more detail without running the significant risk of arriving at non-generalizable conclusions.

Analysis of the Set of Distance Computations.
Review 3 suggests an analysis of which distances are computed in each algorithm. We think this is an idea worth pursuing in the preparation of a full paper, but the actual implementation of this approach seems to require considerable effort.
While obtaining the data needed for such an analysis is straightforward, how to interpret it is still unclear to us at this time. Clusterings on data beyond the two-dimensional case are difficult to interpret visually, making it harder to develop meaningful measures for comparing the distance computation sets.

Minor Problems.
Review 3 correctly notes that Eq. 2 has a typo in the first summand of the RHS, which will be corrected. Likewise, we will also improve the readability and accessibility of Fig. 1 by increasing its font size and using different line styles rather than relying on coloring alone.

### 2025-01-08 review/rebuttal.md

Thank you for your reviews! We greatly appreciate your suggestions and will take them into consideration when updating our paper. Let me address your questions:

Total Runtime Measurements (R2, R3)
Our paper only considers the number of explicit distance evaluations to measure the efficiency of the algorithms we compare. Reviews 2 and 3 bring up the valid point of missing total execution time measurements. While we mention this fact as a limitation (ch. 4.3, Limitations, point ii), our paper does not give a good explanation for this choice.

We chose to count distance evaluations because they are agnostic to the specific implementation of algorithms in machine code. The main disadvantage of reporting execution times is their strong dependence on the amount of optimization that went into this implementation.
For example, time measurements are affected by the programming language chosen, the amount of time spend on optimizing the source code, and the specific choice of hardware architecture (e.g. GPU or CPU). It takes considerable effort and care to make time measurements objective, reproducible, and comparable. Consequently, in our short paper we focus on the theoretical, implementation-independent aspects of the discussed algorithms.

That said, the actual time measurements are important for real-world applications: Besides the distance evaluations, the caching, and evaluation of the upper and lower bound approximations also takes time. While we reduce the total number of distance computations compared to Elkan's algorithm, our new bounds require more memory accesses and computations (cf. Eq. 3,4 and Eq. 5,6).
While this additional time is likely to be small compared to the time spent on distance evaluations, we plan to measure and report the actual ratio in a future full paper.

### 2025-01-08 paper.tex
add Makefile and fix references breaking the entire file
elaborate why we're not using time measurements

\input{content/5_results}

\section{Conclusions and Future Work} \label{sec: conclusions}

% This paper addresses efficient data clustering using the $k$-means algorithm.
% We extend Elkan's algorithm with novel distance approximations derived from Ptolemy's inequality.
% Preliminary evaluations show significant performance improvements over Elkan's method, particularly with large numbers of clusters at low-to-medium dimensionalities.
% The application of Ptolemy's inequality opens new avenues for future work, potentially enhancing other k-means algorithms that use triangle inequality for distance pruning. This approach could make clustering algorithms more practical across a broader range of applications.

This short paper has addressed the problem of efficient data clustering by means of the k-means algorithm.
We have shown how to extend the more efficient clustering algorithm proposed by Elkan with novel distance approximations derived from Ptolemy's inequality.
The results of our preliminary performance evaluation indicate that our proposal achieves significant improvements in performance compared to Elkan's algorithm. %Particularly noteworthy is the increased efficiency with a large number of clusters.

Moreover, the use of the Ptolemy's inequality opens up new avenues for future work.
Besides conducting more extensive experiments to verify our findings,
it would be valuable to explore adaptations of this technique for other k-means algorithms and optimizations.
In particular for algorithms which rely on the triangle inequality to prune distance calculations, Ptolemy's inequality can potentially be leveraged to achieve an enhancement in performance, making these clustering algorithms more practical for a broader range of applications.

% \todo{Moreover, the use of Ptolemy’s inequality opens new avenues for future work. Exploring adaptations of this technique for other K-Means algorithms and optimizations could be highly valuable. In particular, algorithms that rely on the triangle inequality to prune distance calculations might leverage Ptolemy’s inequality to enhance performance, making these clustering algorithms more practical for a broader range of applications.}

\bibliography{references,bibliography_thesis,bibliography_iiwas}
\end{document}

>>>>

\input{content/5_results}
\input{content/6_conclusion.tex}
\bibliography{references,bibliography_thesis,bibliography_iiwas}
\end{document}
